---
title: 'My Daily Travel: mode share by income'
author: "Greta Ritzenthaler"
date: "9/1/2020"
output: html_document
---

Compare My Daily Travel to Travel Tracker for mode share by income groups.

Scroll to bottom for MDT-only analysis.

Cleaning up: 

* removed people less than 16 years old
* removed out-of-region locations
* removed trips greater than 100 miles long
* removed start-of-day "trips" and trips missing modes
* removed observations with missing income (refusal, don't know, etc)


```{r setup, results='hide', echo=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(nord)
library(reldist)
library(knitr)
library(kableExtra)
```

```{r}
#################################################
#                                               #
#                 Data Prep                     #
#                                               #
#################################################

# Load My Daily Travel
setwd("C:/Users/greta/OneDrive - Chicago Metropolitan Agency for Planning/MDT/public")

# trips
trips <- read_csv("place.csv") %>% 
    select(sampno, locno, perno, placeno, placeGroup, mode, distance)

# person info
ppl <- read_csv("person.csv") %>%
    select(sampno, perno, age, hisp, race)

# household info
hh <- read_csv("household.csv") %>% 
    select(sampno, hhinc, hhveh, wthhfin)

# location file w/region flag
region <- read_csv("location.csv") %>% 
    select(sampno, locno, out_region)

# merge datasets
mdt <- trips %>%
    inner_join(ppl, by = c("sampno", "perno")) %>% 
    inner_join(hh, by = "sampno") %>%
    inner_join(region, by = c("sampno", "locno")) %>% 
    # remove out of region
    filter(out_region==0 & distance<=100 & age>=16)

# take care of collapsed trips with placeGroup
mdt <- mdt %>% 
    arrange(desc(distance)) %>% 
    distinct(sampno, perno, placeGroup, .keep_all = TRUE)
    # distinct takes the first row for duplicates, so order by distance to get right mode

rm(list = c("hh", "ppl", "region", "trips"))
```

```{r}
# Load Travel Tracker
# Downloaded from CMAP data portal; exported from Microsoft Access database to csv.
setwd("C:/Users/greta/OneDrive - Chicago Metropolitan Agency for Planning/MDT/traveltracker")

# Household
tt_hh <- read_csv("hh_public.csv") %>%
    select(SAMPN, SURVEY, ASSN, DAY, HHVEH, INCOM, WGTHH)

# people
tt_ppl <- read_csv("per_public.csv") %>%
    select(SAMPN, PERNO, SURVEY, AGE, HISP, RACE, WGTP)

# trips
#   day beginning/place #1 already null in mode var
tt_place <- read_csv("place_public.csv") %>%
    select(MPO, SAMPN, PERNO, DAYNO, PLANO, locno, TPURP, MODE, DIST)

# Combine datasets
#   Remove trips ending outside the region
tt <- tt_place %>%
    inner_join(tt_ppl, by = c("SAMPN", "PERNO")) %>%
    inner_join(tt_hh, by = c("SAMPN", "SURVEY"))

# Flag weekend trips and adjust weights accordingly 
tt <- tt %>%
    mutate(weekend = if_else(SURVEY==2 & DAY==5 & DAYNO==2, 1,
                             if_else(SURVEY==2 & DAY==7 & DAYNO==1, 1, 0)),
           weekdays2 = if_else(SURVEY==2 & DAY==1 | 
                               SURVEY==2 & DAY==2 | 
                               SURVEY==2 & DAY==3 | 
                               SURVEY==2 & DAY==4, 1, 0),
           # If respondent recorded two weekdays, divide weight in half
           weight = if_else(weekdays2==1, WGTP/2, WGTP))

tt <- tt %>%
    filter(MPO==1 & DIST<=100 & weekend==0 & AGE>=16)

rm(list = c("tt_hh", "tt_ppl", "tt_place"))
```


# Income bins

Income groups are different between the two surveys 

```{r echo=TRUE}
tt <- tt %>% 
    mutate(income = factor(INCOM),
           income = recode(income, 
                "1" = "less than $20,000", 
                "2" = "$20,000 to $34,999", 
                "3" = "$35,000 to $49,999", 
                "4" = "$50,000 to $59,999", 
                "5" = "$60,000 to $74,999", 
                "6" = "$75,000 to $99,999", 
                "7" = "$100,000 or more",
                "9" = "refused"))
mdt <- mdt %>% 
    mutate(income = factor(hhinc),
           income = recode(income,
                "1" = "less than $15,000", 
                "2" = "$15,000 to $24,999",
                "3" = "$25,000 to $29,999",
                "4" = "$30,000 to $34,999", 
                "5" = "$35,000 to $49,999", 
                "6" = "$50,000 to $59,999", 
                "7" = "$60,000 to $74,999", 
                "8" = "$75,000 to $99,999", 
                "9" = "$100,000 to $149,999",
                "10" = "$150,000 or more",
                "-9" = "not ascertained",
                "-8" = "don't know", 
                "-7" = "prefer not to answer"))

tt <- tt %>% 
    mutate(income_c = fct_collapse(income,
                low = c("less than $20,000", "$20,000 to $34,999"),
                `middle-low` = c("$35,000 to $49,999", "$50,000 to $59,999"),
                `middle-high` = c("$60,000 to $74,999", "$75,000 to $99,999"),
                high = "$100,000 or more",
                missing = "refused"))

mdt <- mdt %>% 
    mutate(income_c = fct_collapse(income,
                low = c("less than $15,000", "$15,000 to $24,999", "$25,000 to $29,999"),
                `middle-low` = c("$30,000 to $34,999", "$35,000 to $49,999", "$50,000 to $59,999"),
                `middle-high` = c("$60,000 to $74,999", "$75,000 to $99,999"),
                high = c("$100,000 to $149,999", "$150,000 or more"),
                missing = c("not ascertained", "don't know", "prefer not to answer")))
```


# Mode choices

```{r echo=TRUE}
mdt <- mdt %>%
    mutate(mode = factor(mode),
           mode = recode(mode,
                 "101" = "walk",
                 "102" = "personal bike",
                 "103" = "bike share",
                 "104" = "bike share", 
                 "201" = "motorcyle", 
                 "202" = "personal auto (driver)",
                 "203" = "personal auto (passenger)",
                 "301" = "carpool",
                 "401" = "school bus",
                 "500" = "rail and bus",
                 "501" = "bus",
                 "502" = "paratransit", 
                 "503" = "paratransit", 
                 "504" = "paratransit",
                 "505" = "train", 
                 "506" = "local transit",
                 "509" = "transit",
                 "601" = "private shuttle",
                 "701" = "taxi",
                 "702" = "private limo",
                 "703" = "private car",
                 "704" = "rideshare",
                 "705" = "shared rideshare",
                 "801" = "airplane",
                 "997" = "other",
                 "-9"  = "missing",
                 "-1" = "beginning"))

tt <- tt %>% 
    mutate(MODE = factor(MODE),
           MODE = recode(MODE,
                 "1"  = "walk",
                 "2"  = "bike",
                 "3"  = "personal auto (driver)",
                 "4"  = "personal auto (passenger)",
                 "5"  = "CTA bus",
                 "6"  = "CTA train",
                 "7"  = "Pace",
                 "8"  = "Metra",
                 "9"  = "private shuttle",
                 "10" = "paratransit",
                 "11" = "school bus",
                 "12" = "taxi",
                 "14" = "local transit",
                 "15" = "transit (many)",
                 "97" = "other",
                 "98" = "missing",
                 "99" = "missing"))

# condense into mode categories
mdt <- mdt %>%
    mutate(mode_c = fct_collapse(mode,
                     walk = "walk",
                     bike = c("personal bike", "bike share"),
                     transit = c("rail and bus", "bus", "train", "local transit", "transit"),
                     driver = c("motorcyle", "personal auto (driver)"),
                     passenger = c("personal auto (passenger)", "carpool"),
                     other = c("school bus", "paratransit", "private shuttle", 
                               "taxi", "private limo", "private car", "rideshare", 
                               "shared rideshare", "airplane", "other"),
                     missing = "missing", 
                     beginning = "beginning")) %>% 
    # remove missing and beginning modes
    filter(mode_c!="missing" & mode_c!="beginning")

tt <- tt %>%
    mutate(mode_c = fct_collapse(MODE,
                    walk = "walk",
                    bike = "bike",
                    transit = c("CTA bus", "CTA train", "Pace", "Metra", 
                                "local transit", "transit (many)"),
                    driver = "personal auto (driver)",
                    passenger = "personal auto (passenger)",
                    other = c("private shuttle", "paratransit", "school bus", 
                              "taxi", "other"),
                    missing = "missing")) %>% 
    # remove missing and beginning modes
    filter(PLANO!=1 & mode_c!="missing")
```


# Comparing mode share in custom categories between years

* Low:         < 30k (MDT); < 20k & 24-34k (TT)
* Middle-low:  30-59k (MDT); 35-49k & 49-59k (TT)
* Middle-high: 60-99k (MDT); 60-59k & 60-75K & 75-99k (TT)
* High:        everything else

```{r}
inc_total_mdt <- mdt %>%
    filter(income_c!="missing") %>% 
    group_by(income_c) %>%
    summarise(total = sum(wthhfin))

mode_inc_mdt <- mdt %>%
    filter(income_c!="missing") %>% 
    group_by(income_c, mode_c) %>%
    summarise(mode_count = sum(wthhfin)) %>%
    left_join(inc_total_mdt, by = "income_c") %>%
    mutate(mode_share = (mode_count / total) * 100) %>% 
    mutate(mode_c = fct_relevel(mode_c, 
                                levels = c("driver", "passenger", "other", "bike", 
                                           "walk", "transit")))
```

```{r}
inc_total_tt <- tt %>%
    filter(income_c!="missing") %>% 
    group_by(income_c) %>%
    summarise(total = sum(weight))

mode_inc_tt <- tt %>%
    filter(income_c!="missing") %>% 
    group_by(income_c, mode_c) %>%
    summarise(mode_count = sum(weight)) %>%
    left_join(inc_total_tt, by = "income_c") %>%
    mutate(mode_share = (mode_count / total) * 100) %>% 
    mutate(mode_c = fct_relevel(mode_c, 
                                levels = c("driver", "passenger", "other", "bike", 
                                           "walk", "transit")))
```

```{r}
tt_mode_table <- mode_inc_tt %>% 
    select(income_c, mode_c, mode_share) %>% 
    mutate(mode_share = round(mode_share, digits = 1)) %>% 
    pivot_wider(names_from = mode_c, values_from = mode_share) %>% 
    select(income_c, driver, passenger, transit, walk, bike, other)

kable(tt_mode_table, caption = "Travel Tracker") %>% 
    kable_styling(bootstrap_options = "striped")
```

```{r}
mdt_mode_table <- mode_inc_mdt %>% 
    select(income_c, mode_c, mode_share) %>% 
    mutate(mode_share = round(mode_share, digits = 1)) %>% 
    pivot_wider(names_from = mode_c, values_from = mode_share) %>% 
    select(income_c, driver, passenger, transit, walk, bike, other)

kable(mdt_mode_table, caption = "My Daily Travel") %>% 
    kable_styling(bootstrap_options = "striped")
```

```{r}
# matrix calc to find change in percentage points
mode_diff_table <- mdt_mode_table[-1] - tt_mode_table[-1]

mode_diff_table$income_c <- mdt_mode_table$income_c

mode_diff_table <- mode_diff_table %>% 
    select(income_c, driver:other)

kable(mode_diff_table, caption = "Change in percentage points") %>% 
    kable_styling(bootstrap_options = "striped")
```


```{r eval=FALSE}
# lowest
tt1 <- tt %>% 
    filter(income=="less than $20,000" | income=="$20,000 to $34,999") %>% 
    mutate(survey = "Travel Tracker") %>% 
    select(income, survey, weight, mode_c) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

mdt1 <- mdt %>% 
    filter(income=="less than $15,000" | income=="$15,000 to $24,999" | income=="$25,000 to $29,999") %>% 
    mutate(survey = "My Daily Travel") %>% 
    select(income, survey, wthhfin, mode_c) %>% 
    rename(weight = wthhfin) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

low <- rbind(tt1, mdt1)

ggplot(low, aes(x = survey, y = share, fill = mode_c)) +
    geom_col(position = "stack") +
    scale_fill_nord(palette = "red_mountain") +
    coord_flip() +
    labs(title = "Low income")
```


```{r eval=FALSE}
# middle-low
tt2 <- tt %>% 
    filter(income=="$35,000 to $49,999" | income=="$50,000 to $59,999") %>% 
    mutate(survey = "Travel Tracker") %>% 
    select(income, survey, weight, mode_c) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

mdt2 <- mdt %>% 
    filter(income=="$30,000 to $34,999" | income=="$35,000 to $49,999" | income=="$50,000 to $59,999") %>% 
    mutate(survey = "My Daily Travel") %>% 
    select(income, survey, wthhfin, mode_c) %>% 
    rename(weight = wthhfin) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

middlelow <- rbind(tt2, mdt2)

ggplot(middlelow, aes(x = survey, y = share, fill = mode_c)) +
    geom_col(position = "stack") +
    scale_fill_nord(palette = "red_mountain") +
    coord_flip() +
    labs(title = "Middle-low income")
```


```{r eval=FALSE}
# middle-high
tt3 <- tt %>% 
    filter(income=="$60,000 to $74,999" | income=="$75,000 to $99,999") %>% 
    mutate(survey = "Travel Tracker") %>% 
    select(income, survey, weight, mode_c) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

mdt3 <- mdt %>% 
    filter(income=="$60,000 to $74,999" | income=="$75,000 to $99,999") %>% 
    mutate(survey = "My Daily Travel") %>% 
    select(income, survey, wthhfin, mode_c) %>% 
    rename(weight = wthhfin) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

middlehigh <- rbind(tt3, mdt3)

ggplot(middlehigh, aes(x = survey, y = share, fill = mode_c)) +
    geom_col(position = "stack") +
    scale_fill_nord(palette = "red_mountain") +
    coord_flip() +
    labs(title = "Middle-high income")
```


```{r eval=FALSE}
# high
tt4 <- tt %>% 
    filter(income=="$100,000 or more") %>% 
    mutate(survey = "Travel Tracker") %>% 
    select(income, survey, weight, mode_c) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

mdt4 <- mdt %>% 
    filter(income=="$100,000 to $149,999" | income=="$150,000 or more") %>% 
    mutate(survey = "My Daily Travel") %>% 
    select(income, survey, wthhfin, mode_c) %>% 
    rename(weight = wthhfin) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c, survey) %>% 
    summarize(count = sum(weight),
              share = count / median(total))

high <- rbind(tt4, mdt4)

ggplot(high, aes(x = survey, y = share, fill = mode_c)) +
    geom_col(position = "stack") +
    scale_fill_nord(palette = "red_mountain") +
    coord_flip() +
    labs(title = "High income")
```


# Compare all bins of MDT (ignore Travel Tracker)

```{r}
inc_totals_mdt <- mdt %>%
    filter(income_c!="missing") %>% 
    group_by(income) %>%
    summarise(total = sum(wthhfin))

mc_inc_mdt <- mdt %>%
    filter(income_c!="missing") %>% 
    group_by(income, mode_c) %>%
    summarise(mode_count = sum(wthhfin)) %>%
    left_join(inc_totals_mdt, by = "income") %>%
    mutate(mode_share = (mode_count / total) * 100) %>% 
    mutate(mode_c = fct_relevel(mode_c, 
                                levels = c("driver", "passenger", "other", "bike", "walk", "transit")))
```


```{r}
mc_inc_mdt %>% 
    mutate(mode_share = if_else(mode_c=="driver", mode_share*-1, mode_share),
           mode_share = if_else(mode_c=="passenger", mode_share*-1, mode_share)) %>% 
    ggplot(.) +
        geom_col(aes(x = income, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("red_mountain") +
        labs(title = "My Daily Travel") +
        scale_y_continuous(limits = c(-90, 40),
                           breaks = c(-80, -60, -40, -20, 0, 20, 40)) +
        coord_flip()
```


```{r}
mc_inc_mdt %>% 
    select(income, mode_c, mode_share) %>% 
    mutate(mode_share = round(mode_share, digits = 1)) %>% 
    pivot_wider(names_from = mode_c, values_from = mode_share) %>% 
    select(income, driver, passenger, transit, walk, bike, other) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = "striped")

```

