---
title: "My Daily Travel: mode share by age"
author: "Greta Ritzenthaler"
date: "9/1/2020"
output: html_document
---

Find mode type by traveler age and compare to previous travel survey. Mode categories grouped according to former categories used in 2008 Travel Tracker for comparison.


```{r setup, results='hide', echo=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(nord)
library(knitr)
library(kableExtra)
```

```{r}
#################################################
#                                               #
#                 Data Prep                     #
#                                               #
#################################################

# Load My Daily Travel
setwd("C:/Users/greta/OneDrive - Chicago Metropolitan Agency for Planning/MDT/public")

# latest edits as of 8/31

# trips
trips <- read_csv("place.csv") %>% 
    select(sampno, locno, perno, placeno, placeGroup, mode, distance)

# person info
ppl <- read_csv("person.csv") %>%
    select(sampno, perno, age, hisp, race, wtperfin)

# household info
hh <- read_csv("household.csv") %>% 
    select(sampno, hhinc, hhveh, wthhfin)

# location file w/region flag
region <- read_csv("location.csv") %>% 
    select(sampno, locno, out_region)

# merge datasets
mdt <- trips %>%
    inner_join(ppl, by = c("sampno", "perno")) %>% 
    inner_join(hh, by = "sampno") %>% 
    inner_join(region, by = c("sampno", "locno")) %>% 
    filter(out_region==0 & distance<=100)

# take care of collapsed trips with placeGroup
mdt <- mdt %>% 
    arrange(desc(distance)) %>% 
    distinct(sampno, perno, placeGroup, .keep_all = TRUE)
    # distinct takes the first row for duplicates, so order by distance to get right mode

rm(list = c("trips", "ppl", "hh", "region"))
```

```{r}
# Load Travel Tracker
# Downloaded from CMAP data portal; exported from Microsoft Access database to csv.
setwd("C:/Users/greta/OneDrive - Chicago Metropolitan Agency for Planning/MDT/traveltracker")

# Household
tt_hh <- read_csv("hh_public.csv") %>%
    select(SAMPN, SURVEY, ASSN, DAY, HHVEH, INCOM)

# people
tt_ppl <- read_csv("per_public.csv") %>%
    select(SAMPN, PERNO, SURVEY, AGE, HISP, RACE, WGTP)

# trips
#   day beginning/place #1 already null in mode var
tt_place <- read_csv("place_public.csv") %>%
    select(MPO, SAMPN, PERNO, DAYNO, PLANO, locno, TPURP, MODE, DIST)

# Combine datasets
#   Remove trips ending outside the region
tt <- tt_place %>%
    inner_join(tt_ppl, by = c("SAMPN", "PERNO")) %>%
    inner_join(tt_hh, by = c("SAMPN", "SURVEY"))

# Flag weekend trips and adjust weights accordingly 
tt <- tt %>%
    mutate(weekend = if_else(SURVEY==2 & DAY==5 & DAYNO==2, 1,
                             if_else(SURVEY==2 & DAY==7 & DAYNO==1, 1, 0)),
           weekdays2 = if_else(SURVEY==2 & DAY==1 | 
                               SURVEY==2 & DAY==2 | 
                               SURVEY==2 & DAY==3 | 
                               SURVEY==2 & DAY==4, 1, 0),
           # If respondent recorded two weekdays, divide weight in half
           weight = if_else(weekdays2==1, WGTP/2, WGTP))

tt <- tt %>%
    filter(MPO==1 & DIST<=100 & weekend==0)

rm(list = c("tt_place", "tt_ppl", "tt_hh"))
```


# Mode choices

Note differences in "other" categories (specifically rideshare)

```{r echo=TRUE}
# recode mode factors
mdt <- mdt %>%
    mutate(mode = factor(mode),
           mode = recode(mode,
                 "101" = "walk",
                 "102" = "personal bike",
                 "103" = "bike share",
                 "104" = "bike share", 
                 "201" = "motorcyle", 
                 "202" = "personal auto (driver)",
                 "203" = "personal auto (passenger)",
                 "301" = "carpool",
                 "401" = "school bus",
                 "500" = "rail and bus",
                 "501" = "bus",
                 "502" = "paratransit", 
                 "503" = "paratransit", 
                 "504" = "paratransit",
                 "505" = "train", 
                 "506" = "local transit",
                 "509" = "transit",
                 "601" = "private shuttle",
                 "701" = "taxi",
                 "702" = "private limo",
                 "703" = "private car",
                 "704" = "rideshare",
                 "705" = "shared rideshare",
                 "801" = "airplane",
                 "997" = "other",
                 "-9"  = "missing",
                 "-1" = "beginning"))

tt <- tt %>% 
    mutate(MODE = factor(MODE),
           MODE = recode(MODE,
                 "1"  = "walk",
                 "2"  = "bike",
                 "3"  = "personal auto (driver)",
                 "4"  = "personal auto (passenger)",
                 "5"  = "CTA bus",
                 "6"  = "CTA train",
                 "7"  = "Pace",
                 "8"  = "Metra",
                 "9"  = "private shuttle",
                 "10" = "paratransit",
                 "11" = "school bus",
                 "12" = "taxi",
                 "14" = "local transit",
                 "15" = "transit (many)",
                 "97" = "other",
                 "98" = "missing",
                 "99" = "missing"))

# condense into mode categories
mdt <- mdt %>%
    mutate(mode_c = fct_collapse(mode,
                     walk = "walk",
                     bike = c("personal bike", "bike share"),
                     transit = c("rail and bus", "bus", "train", "local transit", "transit"),
                     driver = c("motorcyle", "personal auto (driver)"),
                     passenger = c("personal auto (passenger)", "carpool"),
                     other = c("school bus", "paratransit", "private shuttle", 
                               "taxi", "private limo", "private car", "rideshare", 
                               "shared rideshare", "airplane", "other"),
                     missing = "missing", 
                     beginning = "beginning"))

tt <- tt %>%
    mutate(mode_c = fct_collapse(MODE,
                    walk = "walk",
                    bike = "bike",
                    transit = c("CTA bus", "CTA train", "Pace", "Metra", 
                                "local transit", "transit (many)"),
                    driver = "personal auto (driver)",
                    passenger = "personal auto (passenger)",
                    other = c("private shuttle", "paratransit", "school bus", 
                              "taxi", "other"),
                    missing = "missing"))
```


# Mode share overall

In this data subset:

* ages 5+
* trips out of region removed
* trips over 100 miles long removed
* trips w/missing mode removed

```{r}
mdt.mode.all <- mdt %>% 
    filter(mode_c!="missing" & mode_c!="beginning" & age>=5) %>% 
    mutate(total = sum(wthhfin)) %>% 
    group_by(mode_c) %>% 
    summarise(count = sum(wthhfin),
              mdt_share = round((count/median(total))*100, digits = 2)) %>% 
    select(mode_c, mdt_share)

tt.mode.all <- tt %>% 
    filter(mode_c!="missing" & AGE>=5) %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_c) %>% 
    summarize(count = sum(weight),
              tt_share = round((count/median(total))*100, digits = 2)) %>% 
    select(mode_c, tt_share)
    
mode.all <- tt.mode.all %>% 
    left_join(mdt.mode.all, by = "mode_c") %>% 
    mutate(difference = mdt_share - tt_share)
```

```{r}
kable(mode.all) %>% 
    kable_styling(bootstrap_options = "striped")
```


## Examining the 'other' category for TNC

No TNC category available from Travel Tracker

```{r}
mdt <- mdt %>% 
    mutate(mode_o = fct_collapse(mode,
                    tnc = c("rideshare", "shared rideshare"),
                    taxi = c("taxi"),
                    other = c("school bus", "paratransit", "private shuttle", 
                              "private limo", "private car", "airplane", "other")))

tt <- tt %>% 
    mutate(mode_o = fct_collapse(MODE,
                    taxi = c("taxi"),
                    other = c("private shuttle", "paratransit", "school bus", "other")))

mdt.other <- mdt %>% 
    filter(mode_c=="other") %>% 
    mutate(total = sum(wthhfin)) %>% 
    group_by(mode_o) %>% 
    summarise(mdt_count = sum(wthhfin),
              mdt_share = round((mdt_count/median(total))*100, digits = 2)) %>% 
    mutate(mdt_count = round(mdt_count))

tt.other <- tt %>% 
    filter(mode_c=="other") %>% 
    mutate(total = sum(weight)) %>% 
    group_by(mode_o) %>% 
    summarise(tt_count = sum(weight),
              tt_share = round((tt_count/median(total))*100, digits = 2)) %>% 
    mutate(tt_count = round(tt_count))

other <- tt.other %>% 
    full_join(mdt.other, by = "mode_o")
```

```{r}
kable(other) %>% 
    kable_styling(bootstrap_options = "striped")
```


# Mode share by age

### Subsetting the data

* Remove 0-4 year olds for consistency between datasets
* Age ceiling at 90
* Trips out of region removed
* Trips over 100 miles long removed
* Remove non-responses (-7, -8)
* Remove "beginning" mode

```{r}
# Age bins
decades <- c(-1, 10, 20, 30, 40, 50, 60, 70, 80, 90)
age_labels <- c("5 to 9", "10 to 19", "20 to 29", "30 to 39", "40 to 49", 
                "50 to 59", "60 to 69", "70 to 79", "80 to 89")

mdt <- mdt %>%
    mutate(age_bin = cut(age, breaks = decades, 
                         labels = age_labels))

tt <- tt %>%
    mutate(age_bin = cut(AGE, breaks = decades,
                         labels = age_labels))
```


### My Daily Travel:
```{r}
age_total_mdt <- mdt %>%
    filter(age < 90 & age>=5 & mode_c!="missing"& mode_c!="beginning") %>%
    group_by(age_bin) %>%
    summarise(total = sum(wthhfin))

mc_age_mdt <- mdt %>%
    filter(age < 90 & age>=5 & mode_c!="missing"& mode_c!="beginning") %>%
    group_by(age_bin, mode_c) %>%
    summarise(mode_count = sum(wthhfin)) %>%
    left_join(age_total_mdt, by = "age_bin") %>%
    mutate(mode_share = (mode_count / total) * 100) %>% 
    mutate(mode_c = fct_relevel(mode_c, 
                                levels = c("driver", "passenger", "other", "bike", 
                                           "walk", "transit")))
```

Number of observations for this subset:

* 99485 as of 4/10 (without region flag)
* 97778 as of 4/16
* 97781 as of 5/12
* 97705 as of 5/26 
* 96844 as of 8/27
* 96735 as of 9/1

```{r eval=FALSE}
mdt %>%
    filter(age < 90 & age>=5 & mode_c!="missing" & mode_c!="beginning") %>%
    nrow()
```

### Travel Tracker:
```{r}
age_total_tt <- tt %>%
    filter(AGE < 90 & AGE>=5 & PLANO!=1 & mode_c!="missing") %>%
    group_by(age_bin) %>%
    summarise(total = sum(weight))

mc_age_tt <- tt %>%
    filter(AGE < 90 & AGE>=5 & PLANO!=1 & mode_c!="missing") %>%
    group_by(age_bin, mode_c) %>%
    summarise(mode_count = sum(weight)) %>%
    left_join(age_total_tt, by = "age_bin") %>%
    mutate(mode_share = (mode_count / total) * 100) %>% 
    mutate(mode_c = fct_relevel(mode_c, 
                                levels = c("driver", "passenger", "other", "bike", 
                                           "walk", "transit")))
```

Number of observations for this subset:

* 104181 as of 4/10
```{r eval=FALSE}
tt %>%
    filter(AGE < 90 & AGE>=5 & PLANO!=1 & mode_c!="missing") %>%
    nrow()
```


```{r echo=FALSE}
mc_age_tt %>% 
    mutate(mode_share = if_else(mode_c=="driver", mode_share*-1, mode_share),
           mode_share = if_else(mode_c=="passenger", mode_share*-1, mode_share)) %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("red_mountain") +
        labs(title = "Travel Tracker") +
        scale_y_continuous(limits = c(-90, 40),
                           breaks = c(-80, -60, -40, -20, 0, 20, 40)) +
        coord_flip()


mc_age_mdt %>% 
    mutate(mode_share = if_else(mode_c=="driver", mode_share*-1, mode_share),
           mode_share = if_else(mode_c=="passenger", mode_share*-1, mode_share)) %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("red_mountain") +
        labs(title = "My Daily Travel") +
        scale_y_continuous(limits = c(-90, 40),
                           breaks = c(-80, -60, -40, -20, 0, 20, 40)) +
        coord_flip()
        
```


#### Table format:

```{r}
tt_mode_table <- mc_age_tt %>% 
    select(age_bin, mode_c, mode_share) %>% 
    mutate(mode_share = round(mode_share, digits = 1)) %>% 
    pivot_wider(names_from = mode_c, values_from = mode_share) %>% 
    select(age_bin, driver, passenger, transit, walk, bike, other)

mdt_mode_table <- mc_age_mdt %>% 
    select(age_bin, mode_c, mode_share) %>% 
    mutate(mode_share = round(mode_share, digits = 1)) %>% 
    pivot_wider(names_from = mode_c, values_from = mode_share) %>% 
    select(age_bin, driver, passenger, transit, walk, bike, other)
```


```{r}
kable(tt_mode_table, caption = "Travel tracker") %>% 
    kable_styling(bootstrap_options = "striped")
```


```{r}
kable(mdt_mode_table, caption = "My Daily Travel") %>% 
    kable_styling(bootstrap_options = "striped")
```


```{r}
# matrix calc to find change in percentage points
mode_diff_table <- mdt_mode_table[-1] - tt_mode_table[-1]

mode_diff_table$age_bin <- mdt_mode_table$age_bin

mode_diff_table <- mode_diff_table %>% 
    select(age_bin, driver:other)

kable(mode_diff_table, caption = "Difference in percentage points") %>% 
    kable_styling(bootstrap_options = "striped")
```


#### Close-up of each graph side:

```{r echo=FALSE}
mc_age_tt %>% 
    filter(mode_c=="driver" | mode_c=="passenger") %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("algoma_forest") +
        labs(title = "Travel Tracker, private modes") +
        scale_y_continuous(limits = c(0,90),
                           breaks = c(0, 20, 40, 60, 80)) +
        coord_flip()

mc_age_mdt %>% 
    filter(mode_c=="driver" | mode_c=="passenger") %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("algoma_forest") +
        labs(title = "My Daily Travel, private modes") +
        scale_y_continuous(limits = c(0,90),
                           breaks = c(0, 20, 40, 60, 80)) +
        coord_flip()

mc_age_tt %>% 
    filter(mode_c=="walk" | mode_c=="bike" | mode_c=="transit" | mode_c=="other") %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("lake_superior") +
        labs(title = "Travel Tracker, public modes") +
        coord_flip() +
        scale_y_continuous(limits = c(0, 40))

mc_age_mdt %>% 
    filter(mode_c=="walk" | mode_c=="bike" | mode_c=="transit" | mode_c=="other") %>% 
    ggplot(.) +
        geom_col(aes(x = age_bin, y = mode_share, fill = mode_c),
                 position = "stack") +
        scale_fill_nord("lake_superior") +
        labs(title = "My Daily Travel, public modes") +
        coord_flip() +
        scale_y_continuous(limits = c(0, 40))
```


